\bf\mftitle

### 2011-01-19 doc/jInfer/architecture/architecture.tex
Documentation: some more work on Architecture.

The description of jInfer architecture will commence by describing the data
structures, namely representations of regular expressions and XML elements,
attributes and simple data.

Afterwards the interfaces of basic inference modules - \jmodule{Initial Grammar Generator}, \jmodule{Simplifier} and \jmodule{Schema Generator} - will be explained.

Finally, the process of inference will be described.

@@

\section{Data structures}
\subsection{Regular expressions}
For general information on regular expressions, please refer to \cite{wikiregexp}, \cite{automatatheory}.
All classes pertaining to regular expressions can be found in the package \code{cz.cuni.mff.ksi.jinfer.base.regexp}.
In jInfer, we use extended regular expressions as they give us nicer syntax (and easier programming).

@@

\subsection{Rules and grammars}

@@

\section{Inference process}
\begin{figure}
	\centering\includegraphics[scale=\myscale]{inference_process}
	\caption{High-level view of the inference process} \label{inference_process}
\end{figure}

@@

\subsection{Node metadata} \label{section_metadata}
To allow simple extensibility of rules, class \code{AbstractNamedNode} contains a string-addressed map called \code{metadata} that can contain arbitrary object values.\\

% TODO vektor Describe requirements on what gets passed on between modules

@@

From developer's point of view, inference modules are just properly annotated classes implementing one of the following interfaces (all found under \code{cz.cuni.mff.ksi.jinfer.base.interfaces.inference})

@@

\jmodule{Simplifier} and \jmodule{Schema Generator} take grammar, % TODO vektor Link to grammar definition
in other words a list of \code{Element}s as input. In the first case, this grammar is the Initial Grammar, in the second case it is the simplified grammar.

\jmodule{Initial Grammar Generator} takes an object of type \code{Input}. This class encapsulates all the input files in 3 collections of \code{File}: \code{documents}, \code{schemas} and \code{queries}. Enumerating these files provides IGG with access to all data it needs to create Initial Grammar.\\
\paragraph{Initial Grammar} has two formats:
\begin{itemize}
	\item Only concatenations. This is default format. All regular expressions coming from IGG have to be concatenations, even empty.
	When importing XML documents, this is just fine, as document is positive example from language we are trying to identify.
	But even when importing schemas, queries or other stuff that can contain more information, one has to convert it somehow to concatenations.
	For example, when importing DTD:
	\begin{verbatim}
	<!ELEMENT person (info?, more, more?, more?)>
	<!ELEMENT info (#PCDATA | note)*>
	<!ELEMENT note EMPTY>
	<!ELEMENT more EMPTY>
	\end{verbatim}
	Elements note and more are empty, but we cannot use lambda regexp. Import will use empty concatenation instead.
	For element person, import module have to generate some positive examples of language generated by regular expression \verb+(info?, more, more?, more?)+. This generating is basicly done by calling \code{cz.cuni.mff.ksi.jinfer.base.interfaces.Expander} interface method \code{expand()}. One can implement own intelligent expander, or use our provided \code{cz.cuni.mff.ksi.jinfer.basicigg.expansion.ExpanderImpl}.

@@

@@

\section{NetBeans Modules}

@@

A short overview of these modules and their functions follow. To get a deeper understanding of any of them, please refer to its documentation.\\

@@

\begin{itemize}
	\item \jmodule{jInfer}: not a NBP module in fact, but rather a module suite - the ,,root" NB project for the whole framework.
	\item \jmodule{Base}: contains common data structures, interfaces and utility logic shared across all other modules.
\end{itemize}

@@

\begin{itemize}
	\item \jmodule{RuleDisplayer}: module capable of displaying a grammar in a graphical way.
	\item \jmodule{AdvancedRuleDisplayer}: more advanced version of the former using \jmodule{JUNG}'s graphing capabilities.
	\item \jmodule{AutoEditor}: encapsulation of an interactive finite state automaton editor.
	\item \jmodule{JUNG}: wrapper module for the JUNG (see \cite{jung}).
	\item \jmodule{Options}: NBP specific: this module adds a jInfer category in NetBeans Options window.
	\item \jmodule{ProjectType}: NBP specific: this module defines a jInfer project type, along with its input and output files, settings etc.
	\item \jmodule{XPathFileType}: NBP specific: this module teaches NetBeans to recognize a \code{.xpath} file extension.
\end{itemize}

TODO vektor Lookups

% TODO module dependency hierarchy
% TODO picture showing out inference modules in a humble graph

To get deeper understanding of NBP modules, their versions, dependencies etc, refer to \cite{modules_api}.

>>>>

The description of jInfer architecture will be divided into 3 main views: the data view, the process view and the plaform view.\\

The data view will commence by describing the data structures, namely representations of regular expressions and XML elements, attributes and simple data. The notions of a \emph{rule} and a \emph{grammar} will be explained, along with the finite state automata representation we use.\\

Afterwards the process of inference will be described, both from high-level and programmatic point of view.\\

Finally, we will look at jInfer as at an NetBeans Platform application consisting of a number of modules performing their tasks and communicating together through a set of well-defined interfaces.\\

@@

\section{Data view}
\subsection{Regular expressions}
For general information on regular expressions, please refer to \cite{wikiregexp}, \cite{automatatheory}.
All classes pertaining to regular expressions can be found in the package \code{cz.cuni.mff.ksi.jinfer.base.regexp}.
In jInfer, we use extended regular expressions as they give us nicer syntax (and easier programming).

@@

\subsection{Rules and grammars} \label{rules_grammar}

@@

\section{Process view} \label{section_inference_process}
\begin{figure}
	\centering\includegraphics[scale=\myscale]{inference_process}
	\caption{High-level view of the inference process} \label{inference_process}
\end{figure}

@@

@@

From developer's point of view, inference modules are just properly annotated classes (see \ref{section_lookups}) implementing one of the following interfaces (all found under \code{cz.cuni.mff.ksi.jinfer.base.interfaces.inference})

@@

\jmodule{Simplifier} and \jmodule{Schema Generator} take grammar (see \ref{rules_grammar}), in other words a list of \code{Element}s as input. In the first case, this grammar is the Initial Grammar, in the second case it is the simplified grammar.

\jmodule{Initial Grammar Generator} takes an object of type \code{Input}. This class encapsulates all the input files in 3 collections of \code{File}: \code{documents}, \code{schemas} and \code{queries}. Enumerating these files provides IGG with access to all data it needs to create Initial Grammar.\\

\paragraph{Initial Grammar} has two formats:
\begin{itemize}
	\item Only concatenations. This is default format. All regular expressions coming from IGG have to be concatenations, even empty.
	When importing XML documents, this is just fine, as document is positive example from language we are trying to identify.
	But even when importing schemas, queries or other stuff that can contain more information, one has to convert it somehow to concatenations.
	For example, when importing DTD:
	\begin{verbatim}
	<!ELEMENT person (info?, more, more?, more?)>
	<!ELEMENT info (#PCDATA | note)*>
	<!ELEMENT note EMPTY>
	<!ELEMENT more EMPTY>
	\end{verbatim}
	Elements note and more are empty, but we cannot use lambda regexp. Import will use empty concatenation instead.
	For element person, import module have to generate some positive examples of language generated by regular expression \verb+(info?, more, more?, more?)+. This generating is basicly done by calling \code{cz.cuni.mff.ksi.jinfer.base.interfaces.Expander} interface method \code{expand()}. One can implement own intelligent expander, or use our provided \code{cz.cuni.mff.ksi.jinfer.basicigg.expansion.ExpanderImpl}.

@@

\subsection{Node metadata} \label{section_metadata} % TODO vektor Move after Programmatic view
To allow simple extensibility of rules, class \code{AbstractNamedNode} contains a string-addressed map called \code{metadata} that can contain arbitrary object values.\\

@@

\section{Platform view}

@@

A short overview of these modules and their functions follows. To get a deeper understanding of any of them, please refer to its documentation. To get deeper understanding of how NBP modules work, their versions, dependencies etc, refer to \cite{modules_api}.\\

@@

\begin{itemize}
	\item \jmodule{jInfer}: not a NBP module in fact, but rather a module suite - the ``root" NB project for the whole framework.
	\item \jmodule{Base}: contains common data structures, interfaces and utility logic shared across all other modules.
\end{itemize}

@@

\begin{itemize}
	\item \jmodule{BasicRuleDisplayer}: module capable of displaying a grammar in a graphical way.
	\item \jmodule{TreeRuleDisplayer}: more advanced version of the former using \jmodule{JUNG}'s graphing capabilities.
	\item \jmodule{AutoEditor}: encapsulation of an interactive finite state automaton editor.
	\item \jmodule{JUNG}: wrapper module for the JUNG (see \cite{jung}).
	\item \jmodule{Options}: NBP specific: this module adds a jInfer category in NetBeans Options window.
	\item \jmodule{ProjectType}: NBP specific: this module defines a jInfer project type, along with its input and output files, settings etc.
	\item \jmodule{XPathFileType}: NBP specific: this module teaches NetBeans to recognize a \code{.xpath} file extension.
\end{itemize}

The dependencies between all the modules are summarized in figure \ref{module_deps}. Figure \ref{modules_inference} shows modules implementing the 3 main inference interfaces described in \ref{section_inference_process}.\\

\begin{figure}
	\centering\includegraphics[scale=0.4]{module_deps}
	\caption{Module dependencies} \label{module_deps}
\end{figure}

\begin{figure}
	\centering\includegraphics[scale=0.4]{modules_inference}
	\caption{Inference modules in default jInfer} \label{modules_inference}
\end{figure}

\subsection{Lookups} \label{section_lookups}

\emph{Lookups} are a part of NBP API, a way to achieve late-binding between code based on this platform. An important notion is the one of a \emph{service provider}. This is a class implementing an interface, annotated in the following way (shown on an example from \jmodule{BasicDTDExporter}).
\begin{verbatim}
  @ServiceProvider(service = SchemaGenerator.class)
  public class SchemaGeneratorImpl implements SchemaGenerator {
    ...
\end{verbatim}
The \code{@ServiceProvider} annotation tells NBP that \code{SchemaGeneratorImpl} is to be registered among all providers of the \code{SchemaGenerator} service (or interface).\\

To look up a service provider, or all of them, one can use the \jmodule{Lookups} API in the following way.
\begin{verbatim}
  Collection<SchemaGenerator> generators = Lookup.getDefault().lookupAll(SchemaGenerator.class);

  for (SchemaGenerator g : generators) {
    System.out.println(g.getName());
  }
\end{verbatim}
Call to \code{Lookup.getDefault().lookupAll()} will return a collection of all service providers registered to be providing the \code{SchemaGenerator} interface. This collection can be walked in some way, and on one or more of service providers methods can be called. We didn't have to actually bind providers to logic that uses their services, NB Platform took care of this for us.\\

At this point it is important to understand that when using lookups this way, all service providers are kept as singletons. In case a provider needs to change its state during providing the service (and needs to be ``fresh" each time it is retrieved), it is best to convert this to a factory design pattern: the service provider registered is just a factory returning instances that do the actual work.\\

TODO anti Link to http://bits.netbeans.org/dev/javadoc/org-openide-util-lookup/org/openide/util/Lookup.html as the official documentation to lookup API.

\subsection{Tokens}

TODO vektor
